---
title: "Simulations Pre-Class Project"
date: "Due October 11th, 2017 at 9:00am"
author: "Linde Chen"
output: html_document

---

<style type="text/css">
.table {

    width: 80%;
    margin-left:10%; 
    margin-right:10%;
}
</style>

```{r,setup, echo=FALSE, cache=TRUE}
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 3, digits = 3)
```


#Project Goals:

With this project we will simulate a famous probability problem. This will not require knowledge of probability or statistics but only the logic to follow the steps in order to simulate this problem. This is one way to solve problems by using the computer. 

 1. **Gambler's Ruin**: Suppose you have a bankroll of $1000 and make bets of $100 on a fair game. By simulating the outcome directly for at most 5000 iterations of the game (or hands), estimate:
 
a. the probability that you have "busted" (lost all your money) by the time you have placed your one hundredth bet. 

```{r}
set.seed(0)

GR.bust = function(bankroll, bet, p, hands) {
# returns the hand at which the gambler loses all his money
  
  bust = 0
  while (bankroll >= bet & bust < hands) {
    if (runif(1) > p) {
      bankroll = bankroll + bet
      } else {
      bankroll = bankroll - bet
      } 
    bust = bust + 1
    }
  return(bust)
  }

busted.hand = replicate(5000, GR.bust(1000, 100, 0.5, 5000))
bust.a= length(which(busted.hand < 100)) 
bust.total = 5000

prob.a = bust.a / bust.total
prob.a
```
    
b. the probability that you have busted by the time you have placed your five hundredth bet by simulating the outcome directly. 
    
```{r}
bust.b = length(which(busted.hand < 500))

prob.b = bust.b / 5000
prob.b
```
    
c. the mean time you go bust, given that you go bust within the first 5000 hands.

```{r}
busted.time = busted.hand[which(busted.hand < 5000)]
mean(busted.time)
```
    
d. the mean and variance of your bankroll after 100 hands (including busts).

```{r}
set.seed(0)

GR.bankroll = function(bankroll, bet, p, hands) {
# returns the bankroll held by the gambler at the specified hand
  
  bust = 0
  while (bankroll >= bet & bust < hands) {
    if (runif(1) > p) {
      bankroll = bankroll + bet
      } else {
      bankroll = bankroll - bet
      } 
    bust = bust + 1
    }
  return(bankroll)
}

bankroll.100 = replicate(5000, GR.bankroll(1000, 100, 0.5, 100))

mean(bankroll.100)
var(bankroll.100)
```
    
e. the mean and variance of your bankroll after 500 hands (including busts).

```{r}
bankroll.500 = replicate(5000, GR.bankroll(1000, 100, 0.5, 500))

mean(bankroll.500)
var(bankroll.500)
```
 
Note: you *must* stop playing if your player has gone bust. How will you handle this in the `for` loop?

2. **Markov Chains**. Suppose you have a game where the probability of winning on your first hand is 48%; each time you win, that probability goes up by one percentage point for the next game (to a maximum of 100%, where it must stay), and each time you lose, it goes back down to 48%. Assume you cannot go bust and that the size of your wager is a constant $100.
  
a. Is this a fair game? Simulate one hundred thousand sequential hands to determine the size of your return. Then repeat this simulation 99 more times to get a range of values to calculate the expectation.

```{r}
set.seed(0)

MC.bankroll = function(bet, p, step, hands) {
# returns the bankroll of the gambler after a specified number of hands for a certain wager, given an initial probability and step by which the probability increases after each win
  
  bankroll = 0
  current = 0
  prob = p
  while (current < hands) {
    if (runif(1) < prob) {
      bankroll = bankroll + bet
      prob = prob + step
      if (prob > 1) {
        prob = 1
      }
    } else {
      bankroll = bankroll - bet
      prob = p
    } 
    current = current + 1
  }
  return(bankroll)
}

return = mean(replicate(100, MC.bankroll(100, 0.48, 0.01, 100000)))
return
```
No, this is not a fair game, because the expected return after 10000 hands is negative. In a fair game, the expected return should be the same as the starting bankroll. 

b. Repeat this process but change the starting probability to a new value within 2% either way. Get the expected return after 100 repetitions. Keep exploring until you have a return value that is as fair as you can make it. Can you do this automatically?

```{r}
set.seed(0)

MC.fair.prob = function(increment, init.p) {
# returns the initial probability for which the MC.bankroll function would yield a fair return, given a specified increment of probability increase after each win
  
  mu.return = mean(replicate(100, MC.bankroll(100, init.p, 0.01, 100000)))
  prob = init.p
  while (mu.return < 0) {
    mu.return = MC.bankroll(100, prob+increment, 0.01, 100000)
    prob = prob + increment
  }
  return(prob)
}

return.fair.prob = MC.fair.prob(0.001, 0.48)
return.fair.prob
```
My value for the starting probability which the expected return is as fair as I can make it is about 0.491. 

c. Repeat again, keeping the initial probability at 48%, but this time change the probability increment to a value different from 1%. Get the expected return after 100 repetitions. Keep changing this value until you have a return value that is as fair as you can make it. 

```{r}
set.seed(0)

MC.fair.step = function(init.step, p) {
  
# returns the increment for which the MC.bankroll function would yield a fair return, given a specified initial increment of probability increase after each win and a certain probability
  
  mu.return = mean(replicate(100, MC.bankroll(100, p, init.step, 100000)))
  increment = init.step
  while (mu.return < 0) {
    mu.return = MC.bankroll(100, p, increment+0.001, 100000)
    increment = increment + 0.001
  }
  return(increment)
}

return.fair.step = MC.fair.step(0.01, 0.48)
return.fair.step
```
My value for the increment adjustment which the expected return is as fair as I can make it is about +0.015.